---
title: "STA9750TermProject"
author: "Shawn Meng, Rongnan He, Szufan Chen, Jiarui Guo"
date: "`r format(Sys.time(), '%B %d %Y')`"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, dev = "png", dpi = 300)
library(tidyverse)
library(lubridate)
library(modelr)
library(broom)
library(randomForest)
```
## Background

In this project, we are going to analyze the impact of the NY COVID-19 shutdown to the levels of air pollutants. Around the world, there are reports showed that the shutdown improved the air quality for some regions.
The EPA(United States environmental protection agency) is tracking five common pollutants nationwide. In this project, we will focus on PM2.5 and nitrogen oxide which are closely related to the traffic. Our data was access from [EPA website](https://www.epa.gov/outdoor-air-quality-data/download-daily-data). 
We will limit our analysis on PM2.5 and NO2 because these two pollutants are closely related to the traffic.

## Objective
1. We want to see if there is a significant impact of the NY PAUSE executive order from governor Cuomo on the air pollutant levels of PM2.5 and NO2 in NYC.
2. We are also interested in the long term trend and seasonality influence of the air pollutants in NYC.

## Exploratory data analysis

### Introduction to the dataset
We obtained the data of PM2.5 and NO2 from 2004 to 2020.
```{r dataimport, echo=FALSE, include=FALSE}
read_data <- function(pollutant, state="NY", year = 2004:2020) {
  datamerge <- data.frame()
  for (i in year){
    temp <- read_csv(file = paste0("dataset/", as.character(pollutant), "_", as.character(state), "_", i,".csv" ))
    datamerge <- rbind(datamerge, temp)
  }
  return(as_tibble(datamerge))
}

PM2.5_NY_2004to2020 <- read_data(pollutant="PM25")
#193975 obs of 20 variables
NO2_NY_2004to2020 <- read_data(pollutant="NO2")
#42814 obs of 20 variables
```

```{r data cleaning, echo=FALSE}
##step 2, cleaning and manipulating data
#checking for missing values
PM2.5_NY_2004to2020 %>% select(`Daily Mean PM2.5 Concentration`) %>%
  filter(is.na(`Daily Mean PM2.5 Concentration`))
#no missing value for Daily mean PM2.5
NO2_NY_2004to2020 %>% select(`Daily Max 1-hour NO2 Concentration`) %>%
  filter(is.na(`Daily Max 1-hour NO2 Concentration`))
#no mission value for Daily Max No2
#somehow there are daily mean pm2.5 readings which are negative, these reading probably caused by faulty sensors. 
#we will exclude from the dataset
PM2.5_NY_2004to2020_cleaned <- PM2.5_NY_2004to2020 %>% filter(`Daily Mean PM2.5 Concentration`>=0)
#check for the percentage of the cleaned data to the total data
paste0(format(nrow(PM2.5_NY_2004to2020_cleaned)/nrow(PM2.5_NY_2004to2020)*100,trim=TRUE, digits = 5),"%")
#99.717% of the total data which I think it is reasonable.

#similarly, clean the data for NO2
NO2_NY_2004to2020_cleaned <- NO2_NY_2004to2020 %>%  filter(`Daily Max 1-hour NO2 Concentration`>0)
paste0(format(nrow(NO2_NY_2004to2020_cleaned)/nrow(NO2_NY_2004to2020)*100, trim = TRUE, digits = 5), "%")
#99.615% of total data

#transform the date to the format of datetime 
PM2.5_NY_2004to2020_cleaned <- PM2.5_NY_2004to2020_cleaned %>% 
  mutate(Date = mdy(Date)) %>% mutate(month=month(Date))
NO2_NY_2004to2020_cleaned <- NO2_NY_2004to2020_cleaned %>% 
  mutate(Date = mdy(Date)) %>% mutate(month=month(Date))
# check how many counties involved in this dataset
unique(PM2.5_NY_2004to2020_cleaned$COUNTY)
unique(NO2_NY_2004to2020_cleaned$COUNTY)
# create NYC counties
NYC_counties <- c("Kings", "Bronx", "Queens", "New York", "Richmond")
#creat the date that NY Pause order go into effect
NY_pause <- mdy_h("March 22 2020 8PM", tz="EST")

```
### first look of the NY PAUSE executive order impact


```{r NYC2020, echo=FALSE}
##step 3, exploratory data analysis
#first look at how the shutdown affects the pollutant in 2020 in NYC
#pm2.5 trend in 2020 in NYC

PM2.5_NY_2020 <- PM2.5_NY_2004to2020_cleaned %>% filter(year(Date)==2020)
PM2.5_NY_2020 %>%  select(Date, `Daily Mean PM2.5 Concentration`, COUNTY) %>%  
  filter(COUNTY %in% NYC_counties) %>% 
  ggplot(aes(x = Date, y= `Daily Mean PM2.5 Concentration`, color = COUNTY))+geom_point(alpha=0.3)+
  geom_vline(mapping = aes(xintercept = date(NY_pause))) +geom_smooth(se=FALSE)+
  geom_text(aes(x = date(NY_pause), y = 20, label = "NY PAUSE executive order"), size = 3, hjust=-0.05, color = "black")+
  labs(title = "Daily PM 2.5 concetration in NYC in 2020", x = "", y="Daily Mean PM2.5 concentration (ug/m3 LC)")

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
